---
title: "GCMP Heat Map for Publication R Notebook"
output: html_notebook
---

The following code is for running a heat map against the following:
1) %disease
2) growth rate
3) microbial alpha div
4) microbial dominance
5) endozoicomonas dominance


First load libraries
```{r}
library(phytools)
library(RColorBrewer)
```


Load in a tree file and trait table and filter the trait table into the columns intended for the heat map

```{r}
tree <- read.tree('../input/huang_roy_genus_tree.newick')
traits <- read.csv('../input/GCMP_trait_table.csv', header = TRUE)
heatmap.traits <- traits[,c()] 


```


Set the colors for the heat map
```{r}
disease_color <- c("lightcoral","deepskyblue")
disease_ramp_color_palette <- colorRampPalette(c("lemonchiffon1", "red"))(100)
```


Run the heat map codecode
```{r}
pdf(file="../output/heatmap_standardized.pdf")
phylo.heatmap(ctree, disease_table, standardize=TRUE, fsize=0.6, colors=disease_ramp_color_palette,length=1)
dev.off()
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

